\large How resource abundance and stochasticity affect organisms' range sizes

### 2024-06-06 writing/appendix-A-concepts-and-definitions.Rmd
cleaned up appendix A

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero from the positive side, the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant.

@@

There are many distributions we can assign to $R$ depending on how we quantify it. For instance, if we measure $R$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework). Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The expectation of a random variable is the value one can expect the random variable to take, on average, in the long term. Here we use $\e(R)$ to indicate the expectation of the random variable $R$. When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$ (Fig. A\@ref(fig:multipanel-constant-var)). We indicate the estimated average amount of resources over time by adding a caret on the symbol: $\hat \mu(t)$. Note, however, that the estimate of $\mu(t)$ and what values it can take depend on what distribution we assume $R$ follows. For a Gamma distribution, $\mu(t)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t)$ is necessarily within the interval $(-1, 1)$. See the section on modeling $R$ in the main manuscript for more information on how to choose probability distributions.

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (A) Resources ($R$, orange fill) vary over time and space with a constant the variance (dot color). (B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

@@

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's variance is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$; see Fig. A\@ref(fig:multipanel-changing-var)). <!-- Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others due to changes in $\sigma^2(t)$. Additionally, we distinguish between two different sources of variation. We refer to known or expected spatiotemporal differences or changes as heterogeneity, which we consider as a change in $\mu(t)$. In contrast, we use the term stochasticity for unpredictable variation (i.e., due to $\sigma^2(t)$).--> For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons. Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t)$ to be lower in the winter since it does not expect to find berries (i.e., $\mu(t) \approx 0$). Also note that, like with $\mu(t)$, the estimate of $\sigma^2(t)$ and what values it can take depend on what distribution we assume $R$ follows. Figure A\@ref(fig:ex) shows four examples of Gamma distributions with different means and variances.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (A) Although resources ($R$, orange fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

>>>>

In our simulations, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$ (see the sections below), which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations within the context of resource abundance and stochasticity. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero from the positive side, the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant.

@@

There are many distributions we can assign to $R$ depending on how we quantify it. For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the main text and Appendix C). If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what values we believe $R$ will take, on average, in each of the two locations or time periods. The expectation of a random variable is the value one can expect the random variable to take, on average, over the long term. Here we use $\e(R)$ to indicate the expectation of the random variable $R$. When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$ (Fig. A\@ref(fig:multipanel-constant-var)). If $\e(R)$ changes over time and space, we write $\e(R|t, u) = \mu(t, u)$. We indicate the estimated $\mu(t, u)$ by adding a caret on the symbol: $\hat \mu(t, u)$. Note, however, that the estimate of $\mu(t, u)$ and what values it can take depend on what distribution we assume $R$ follows. For a Gamma distribution, $\mu(t, u)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t, u)$ is necessarily within the interval $(-1, 1)$.

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance around the mean. (A) Resources ($R$, orange fill) vary over time and space with a time-varying but spatially homogeneous mean ($\\mu(t)$) and constant variance (dot color). (B) $R$ follows a Gamma distribution with mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

@@

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$, which we quantify with its variance. We use the notation $\var(R)$ to indicate the variance in $R$ around the mean, i.e., after accounting for changes in $\mu(t, u)$, and we use $\sigma^2(t, u)$ to indicate its function over time and time (with estimate $\widehat{\sigma^2}(t, u)$; see Fig. A\@ref(fig:multipanel-changing-var)). For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons. Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t, u)$ to be lower in the winter since berries will be hard to find (i.e., $\mu(t, u) \approx 0$). Also note that, like with $\mu(t, u)$, what values $\sigma^2(t, u)$ can take depends on what distribution we assume $R$ follows. Figures A\@ref(fig:ex-gamma) and A\@ref(fig:ex-beta) show four examples of Gamma and Beta distributions, respectively, with different means and variances.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance around the mean. (A) Although resources ($R$, orange fill) vary over time and space, the variance (dot color) is lowest at the beginning and end of the observational period, and it is highest when $\\mu(t)$ peaks. (B) $R$ follows a Gamma distribution with time-varying but spatially homogenous mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```
